---
title: "clustering UN nations"
author: "Evan James Martin"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) #, eval = FALSE)
```

## unsupervised learning, clustering and K-means

Unsupervised learning methods have no response variable. The objective is
not related to prediction but rather
to the identification of different possible structures that may be
present in the data. For example, one may be
interested in determining whether the observations 
are grouped in some way (clustering). In this study project we look at clustering nations in the UN. 

The objective of clustering
can be described as identifying different groups 
of observations that are closer to each other than to those of 
the other groups. In general, the number of groups is
unknown and needs to be determined from the data.

Probably the most intuitive and easier to explain
unsupervised clustering algorithm is K-means. We will use K-means on the UN-votes data. Note that the UN-votes data is strictly categorical so using K-means is a bit weird but it turns out it works okay for this example. 

## UN votes example

These data contain the historical voting patterns
of United Nations members. 
The data include only "important" votes, as classified
by the U.S. State Department. The votes for each country 
were coded as follows: Yes (1), Abstain (2), No (3),
Absent (8), Not a Member (9). There were 
368 important votes, and 77 countries
voted in at least 95% of these. We focus on these
UN members. 

Lets explore whether
voting patterns reflect political
alignments, and also whether countries vote along known
political blocks. Our data consists of 77 observations
with 368 variables each. 

We read in the data, and limit ourselves to resolutions where
every country voted without missing votes:
```{r unvotes, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
X <- read.table(file='unvotes.csv', sep=',', row.names=1, header=TRUE)
X2 <- X[complete.cases(X),]
```
We now compute a K-means partition using the function `kmeans`
with *K = 5*, and look at the resulting groups:
```{r kmeans.un1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
set.seed(123)
b <- kmeans(t(X2), centers=5, iter.max=20, nstart=1)
table(b$cluster)
```
Note that if we run `kmeans` again, we might get a different partition:
```{r kmeans.un2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
b <- kmeans(t(X2), centers=5, iter.max=20, nstart=1)
table(b$cluster)
```
This is because K-means is not a determinisitic algorithm and the final clusterings it chooses will depend on an initial random seed. To avoid clusterings that result from an "unlucky" seed, it is better to consider a large number of random starts and take the **best** found solution.
```{r kmeans.un3, fig.width=6, fig.height=6, message=FALSE, warning=FALSE, tidy=TRUE}
# Take the best solution out of 1000 random starts
b <- kmeans(t(X2), centers=5, iter.max=20, nstart=1000)
split(colnames(X2), b$cluster)
```
Lets view the clustering on a map:
```{r kmeans.map1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE, tidy=TRUE}
library(rworldmap)
library(countrycode)
these <- countrycode(colnames(X2), 'country.name', 'iso3c')
malDF <- data.frame(country = these, cluster = b$cluster)
malMap <- joinCountryData2Map(malDF, joinCode = "ISO3",
                              nameJoinColumn = "country")
par(mai=c(0,0,0,0),xaxs="i",yaxs="i")
mapCountryData(malMap, nameColumnToPlot="cluster", catMethod = "categorical",
               missingCountryCol = 'white', addLegend=FALSE, mapTitle="",
               colourPalette=c('darkgreen', 'hotpink', 'tomato', 'blueviolet', 'yellow'),
               oceanCol='dodgerblue')
```

Now we use K-means with K=3 to obtain clusters that may be easier to interpret:

```{r un.pam.map, fig.width=6, fig.height=6, message=FALSE, warning=FALSE, tidy=TRUE}
library(cluster)
these <- countrycode(colnames(X2), 'country.name', 'iso3c')
malDF <- data.frame(country = these, cluster = b$cluster)
malMap <- joinCountryData2Map(malDF, joinCode = "ISO3", nameJoinColumn = "country")
par(mai=c(0,0,0,0),xaxs="i",yaxs="i")
mapCountryData(malMap, nameColumnToPlot="cluster", catMethod = "categorical",
               missingCountryCol = 'white', addLegend=FALSE, mapTitle="",
               colourPalette=c('yellow', 'tomato', 'blueviolet'),
               oceanCol='dodgerblue')
```
